Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories

### 2024-12-18 paper/paper.tex
Moved a comment from 9.2 to 9.1
Added a paragraph about shallow embeddings

More abstractly, Coquand's proof technique is an instance of a categorical
construction called Artin gluing~\citep{kaposi:gluing}. Different
metatheoretic properties, such as normalization, parametricity, and
$\Pi$-injectivity, can be obtained by gluing the initial CwF with the
corresponding CwF model. \citet{sterling:tait} introduces the notion of
synthetic Tait computability, which allows further
simplification of such proofs by working in the internal language
formed by gluing.
However, proof-relevant logical relations are more demanding on the
expressive power of the metatheory. For example, the mechanization of
NbE by \citet{altenkirch:normalisation} relies on quotient
inductive--inductive types (QIIT)~\citep{kaposi:qiit}.

@@

\jc{Missing from the table:
\begin{itemize}
  \item Altenkirch and Kaposi (2016) using QIITs, mentioned above (Agda);
    mechanization might be part of Kaposi's thesis (2016)
  \item prior work Chapman (2009), Type Theory should Eat Itself (Agda);
    possibly also his thesis (2008)
  \item Kipling by McBride (2010), mixing deep and shallow embedding
    using host conversion (Agda) to avoid the whole QIIT business
  \item Kaposi et al. (2019), Shallow Embedding of Type Theory (Agda),
    although at this point I'm not sure if this counts as having a syntax
\end{itemize}
}

Figure~\ref{fig:featurematrix} presents several mechanized proofs
that feature logical relations for dependently-typed languages.
Each of these proofs is significantly larger than our development, but they
also prove more results about different object languages.
The table provides a comparison between the various features of their
object languages, but is not exhaustive. For example,
\citet{casinghino:combining-proofs-programs} and
\citet{anand2014towards} both have support for partial
programs. However, we include features that we believe to be most
impactful to the definition of the logical relation.
For a basic comparison of development size, we include the approximate
nonblank, noncomment lines of proof code for each project.

@@

\citet{decagda} mechanize in Agda the decidability of type
conversion rule for a dependently typed language with one predicative
universe level and a typed judgmental equality that includes the function
$\eta$ law. They
use a Kripke-style logical relation parameterized over a
type-directed equivalence relation satisfying certain
properties to facilitate the reuse of their definition. The
logical relation is defined using induction--recursion,
which is available in Agda but not in Coq. Their development involves
around 10,000 lines of Agda code.
\citet{martin-lof-a-la-coq} adapts the logical relation
by \citet{decagda} to the predicative fragment of Coq and further
extends the decidability of type conversion result to the decidability of
type checking a bidirectional type system.

\citet{anand2014towards} mechanize the metatheory of
Nuprl~\citep{constable1986implementing} in Coq. The object theory is an
extensional type theory with dependent functions, inductive types,
partial types, and a full universe hierarchy. They construct a PER
model in Coq to show the logical consistency of the language. Their
development has been further extended with intersection types,
union types, and quotient types, resulting in a Coq development of around
330,000 lines of code.
\citet{nbeincoq} mechanize a normalization-by-evaluation algorithm in Coq for a
dependently typed language with one predicative universe, similar to
\citet{decagda} and \citet{martin-lof-a-la-coq}. However, since their
type system has no elimination form for natural numbers, the only base type,
large elimination is not supported.
Both \citet{anand2014towards} and \citet{nbeincoq} leverage
Coq's impredicative \texttt{Prop} sort to define the interpretation
of dependent function types and thus are closely related to our
mechanization. \citet{anand2014towards} further show it possible
to encode a finite universe hierarchy using neither impredicativity
nor induction--recursion. Their encoding of a countable
universe hierarchy relies on impredicativity, similar to our
development.
\jc{Wait I'm confused, do they use impredicativity or not?}

>>>>

More abstractly, Coquand's proof technique is an instance of a categorical
construction called Artin gluing~\citep{kaposi:gluing}. Different
metatheoretic properties, such as normalization, parametricity, and
$\Pi$-injectivity, can be obtained by gluing the initial CwF with the
corresponding CwF model. \citet{sterling:tait} introduces the notion of
synthetic Tait computability, which allows further
simplification of such proofs by working in the internal language
formed by gluing.
However, proof-relevant logical relations are more demanding on the
expressive power of the metatheory. For example, the mechanization of
NbE by \citet{altenkirch:normalisation} relies on quotient
inductive--inductive types (QIIT)~\citep{kaposi:qiit}.
\jc{I don't understand how the QIIT model relates to gluing,
or what using QIIT has to do with proof relevance of the logical relation.}

So far, the works mentioned interpret types as sets or relations on deep
representations of terms. In contrast, shallow embeddings interpret types as
sets of \emph{meta-level} terms in some type theory. The logical relation
interprets the object language into the meta language, and its soundness holds
by virtue of the interpretation function being well typed and definable.
Examples include work by \citet{kipling}, who mixes deep and shallow
representations, and work by \citet{shallow}, who directly implement the
signature of a CwF as Agda terms. The key property of shallow embeddings is
that object-level conversion is represented as meta-level equality, which has
the benefit of delegating handling conversion to the type checker, but with the
drawback that object conversion must coincide with meta equality. Conversion
cannot be customized to convert more terms than dictated by their
interpretations.

@@

Figure~\ref{fig:featurematrix} presents several mechanized proofs
that feature logical relations for dependently typed languages.
Each of these proofs is significantly larger than our development, but they
also prove more results about different object languages.
The table provides a comparison between the various features of their
object languages, but is not exhaustive. For example,
\citet{casinghino:combining-proofs-programs} and
\citet{anand2014towards} both have support for partial
programs. However, we include features that we believe to be most
impactful to the definition of the logical relation.
For a basic comparison of development size, we include the approximate
nonblank, noncomment lines of proof code for each project.

@@

\citet{decagda} mechanize in Agda the decidability of type
conversion rule for a dependently typed language with one predicative
universe level and a typed judgmental equality that includes the function
$\eta$ law. They
use a Kripke-style logical relation parameterized over a
type-directed equivalence relation satisfying certain
properties to facilitate the reuse of their definition. The
logical relation is defined using induction--recursion,
which is available in Agda but not in Coq.
\citet{martin-lof-a-la-coq} adapts the logical relation
by \citet{decagda} to the predicative fragment of Coq and further
extends the decidability of type conversion result to the decidability of
type checking a bidirectional type system.

\citet{anand2014towards} mechanize the metatheory of
Nuprl~\citep{constable1986implementing} in Coq. The object theory is an
extensional type theory with dependent functions, inductive types,
partial types, and a full universe hierarchy. They construct a PER
model in Coq to show the logical consistency of the language. Their
development has been further extended with intersection types,
union types, and quotient types.
\citet{nbeincoq} mechanize a normalization-by-evaluation algorithm in Coq for a
dependently typed language with one predicative universe, similar to
\citet{decagda} and \citet{martin-lof-a-la-coq}. However, since their
type system has no elimination form for natural numbers, the only base type,
large elimination is not supported.
Both \citet{anand2014towards} and \citet{nbeincoq} leverage
Coq's impredicative \texttt{Prop} sort to define the interpretation
of dependent function types and thus are closely related to our
mechanization. \citet{anand2014towards} further show it possible
to encode a finite universe hierarchy using neither impredicativity
nor induction--recursion. Their encoding of a countable
universe hierarchy relies on impredicativity, similar to our
development.
\jc{Wait I'm confused, do they use impredicativity or not?}
